Double Elimination Tournament

### 2025-05-01 double_elimination.qmd
update 5/1/25

        wins_team1 <- 0
        wins_team2 <- 0
        losses_team1 <- 0
        losses_team2 <- 0
        for (game in 1:series){
          match_winner <- simulate_match(team1, team2, strength1, strength2)
          #cat(sprintf("Matchup:  %s vs %s | Strengths: %.2f vs %.2f | Winner: %s\n", team1, team2, strength1, strength2, match_winner))
          if(!is.na(team1)) {
          if (match_winner == team1) {
            wins_team1 <- wins_team1 + 1
            losses_team2 <- losses_team2 + 1
          } else {
            wins_team2 <- wins_team2 + 1
            losses_team1 <- losses_team1 + 1
          }
          }
          else {wins_team2 <- (series+1) / 2}

          # Determine the overall winner if a majority is reached
          if (wins_team1 > series / 2 || wins_team2 > series / 2) {
            break
          }
        }
        # Update the game_wins and game_losses outside of the loop
        if (!any(is.na(c(team1, team2)))) {
          #cat(sprintf("Losses - %s: %d | %s: %d\n", team1, losses_team1, team2, losses_team2))
        permuted_df <- permuted_df %>%
          mutate(game_wins = case_when(
            true_rank == team1 ~ game_wins + wins_team1,
            true_rank == team2 ~ game_wins + wins_team2,
            TRUE ~ game_wins
          )) %>%
          mutate(game_losses = case_when(
            true_rank == team1 ~ game_losses + losses_team1,
            true_rank == team2 ~ game_losses + losses_team2,
            TRUE ~ game_losses
          ))
        }
        # Determine the match winner based on series results
        series_winner <- ifelse(wins_team1 > wins_team2, team1, team2)
        series_loser <- ifelse(series_winner == team1, team2, team1)
        #cat(sprintf("Matchup:  %s vs %s | Winner: %s\n: | Loser: %s\n", team1, team2, series_winner, series_loser))
        next_winners <- c(next_winners, series_winner)
        round_losers <- c(round_losers, series_loser)
    }
     winners_bracket <- next_winners  # Winners move to the next round
     losers_bracket <- c(losers_bracket, round_losers)
     if(length(losers_bracket) %% 2 != 0){losers_bracket <- c(losers_bracket, NA)}

@@

    # Losers' bracket
    if (length(losers_bracket) > 0) {
      if(length(losers_bracket) > 1){
      #cat("\n--- Loser's Bracket --- \n")
      }
      next_losers <- c()

@@

          # Determine overall series winner
          if (wins_team1 > series / 2 || wins_team2 > series / 2) {
            break
          }
        }
        if (!any(is.na(c(team1, team2)))) {
          #cat(sprintf("Losses - %s: %d | %s: %d\n", team1, losses_team1, team2, losses_team2))
          permuted_df <- permuted_df %>%
          mutate(game_wins = case_when(
            true_rank == team1 ~ game_wins + wins_team1,
            true_rank == team2 ~ game_wins + wins_team2,
            TRUE ~ game_wins
          )) %>%
          mutate(game_losses = case_when(
            true_rank == team1 ~ game_losses + losses_team1,
            true_rank == team2 ~ game_losses + losses_team2,
            TRUE ~ game_losses
          ))
        }

        if(is.na(team1) && is.na(team2)){
          series_winner <- NA
          next_losers <- c(next_losers, NA)
        } else{
        series_winner <- ifelse(wins_team1 > wins_team2, team1, team2)
        series_loser <- ifelse(wins_team1 > wins_team2, team2, team1)
        next_losers <- c(next_losers, series_winner)
        eliminated <- c(eliminated, series_loser)
        #cat(sprintf("Matchup:  %s vs %s | Winner: %s\n:", team1, team2, series_winner))
        }
      }
        team_rank <- team_rank - 1
        permuted_df$rank_hat[permuted_df$true_rank %in% eliminated] <- team_rank
        losers_bracket <- next_losers

@@

# Final matchup between winners and losers
    #cat("\n--- Final Match ---\n")
    final_match <- c(winners_bracket, losers_bracket)
    if(length(final_match) == 2){
    team1 <- final_match[1]
    team2 <- final_match[2]
    strength1 <- strengths[which(permuted_df$true_rank == team1)]
    strength2 <- strengths[which(permuted_df$true_rank == team2)]

@@

      if (wins_team1 > series / 2 || wins_team2 > series / 2) {
        break
      }
    }
    if (!any(is.na(c(team1, team2)))) {
      #cat(sprintf("Losses - %s: %d | %s: %d\n", team1, losses_team1, team2, losses_team2))
      permuted_df <- permuted_df %>%
        mutate(game_wins = case_when(
          true_rank == team1 ~ game_wins + wins_team1,
          true_rank == team2 ~ game_wins + wins_team2,
          TRUE ~ game_wins
        )) %>%
        mutate(game_losses = case_when(
          true_rank == team1 ~ game_losses + losses_team1,
          true_rank == team2 ~ game_losses + losses_team2,
          TRUE ~ game_losses
          ))
    }

    if (wins_team1 > wins_team2) {
      #cat(sprintf("Matchup:  %s vs %s | Winner: %s\n:", team1, team2, team1))
      permuted_df$rank_hat[permuted_df$true_rank == team1] <- 1
      permuted_df$rank_hat[permuted_df$true_rank == team2] <- 2
    } else {
      #cat(sprintf("Matchup:  %s vs %s | Winner: %s\n:", team1, team2, team2))
      # second game if the loser's bracket team wins
      for (game in 1:series) {
      match_winner <- simulate_match(team1, team2, strength1, strength2)
      if (match_winner == team1) {
        wins_team1 <- wins_team1 + 1
        losses_team2 <- losses_team2 + 1
      } else {
        wins_team2 <- wins_team2 + 1
        losses_team1 <- losses_team1 + 1
      }

      if (wins_team1 > series / 2 || wins_team2 > series / 2) {
        break
      }
    }
      if (!any(is.na(c(team1, team2)))) {
        #cat(sprintf("Wins - %s: %d | %s: %d\n", team1, wins_team1, team2, wins_team2))
        permuted_df <- permuted_df %>%
          mutate(game_wins = case_when(
            true_rank == team1 ~ game_wins + wins_team1,
            true_rank == team2 ~ game_wins + wins_team2 - 1, # wins_team2 = 2 in loop but only need to add one game
            TRUE ~ game_wins
          )) %>%
          mutate(game_losses = case_when(
            true_rank == team1 ~ game_losses + losses_team1 - 1,
            true_rank == team2 ~ game_losses + losses_team2,
            TRUE ~ game_losses
          ))
      }
      if (wins_team1 > wins_team2) {
      #cat(sprintf("Matchup:  %s vs %s | Winner: %s\n:", team1, team2, team1))
      permuted_df$rank_hat[permuted_df$true_rank == team1] <- 1
      permuted_df$rank_hat[permuted_df$true_rank == team2] <- 2
    } else {
      #cat(sprintf("Matchup:  %s vs %s | Winner: %s\n:", team1, team2, team2))
      permuted_df$rank_hat[permuted_df$true_rank == team2] <- 1
      permuted_df$rank_hat[permuted_df$true_rank == team1] <- 2
        }
      }
    }
  }
}
    permuted_df$seed <- rep(paste(seeding_structure[i, ], collapse = ", "), nrow(permuted_df))
    results[[i]] <- permuted_df
}
  final_results <- do.call(rbind, results)

  # Remove ties if ties = FALSE
  if (!ties) {
    final_results <- final_results %>%
      filter(!is.na(true_rank)) %>%
      arrange(true_rank) %>%
      mutate(rank_hat = rank(rank_hat, ties.method = "first"))
  }
  else{
    final_results <- final_results %>%
      filter(!is.na(true_rank)) %>%
      arrange(rank_hat) %>%
      mutate(rank_hat = rank(rank_hat, ties.method = "average"))
  }

@@

#test <- simulate_double_elimination_tournament(num_teams = 8, distribution = "Normal", ties=F, series=1)
#test
#manual_test <- simulate_double_elimination_tournament(num_teams = 4, distribution = "Manual", ties=F, series=1)
#manual_test

>>>>

        wins_team1 <- 0
        wins_team2 <- 0
        losses_team1 <- 0
        losses_team2 <- 0
        for (game in 1:series){
          match_winner <- simulate_match(team1, team2, strength1, strength2)
          cat(sprintf("Matchup:  %s vs %s | Strengths: %.2f vs %.2f | Winner: %s\n", team1, team2, strength1, strength2, match_winner))
          if(!is.na(team1)) {
          if (match_winner == team1) {
            wins_team1 <- wins_team1 + 1
            losses_team2 <- losses_team2 + 1
          } else {
            wins_team2 <- wins_team2 + 1
            losses_team1 <- losses_team1 + 1
          }
          }
          else {wins_team2 <- (series+1) / 2}

          # Determine the overall winner if a majority is reached
          if (wins_team1 > series / 2 || wins_team2 > series / 2) {
            break
          }
        }
        # Update the game_wins and game_losses outside of the loop
        if (!any(is.na(c(team1, team2)))) {
          cat(sprintf("Losses - %s: %d | %s: %d\n", team1, losses_team1, team2, losses_team2))
        permuted_df <- permuted_df %>%
          mutate(game_wins = case_when(
            true_rank == team1 ~ game_wins + wins_team1,
            true_rank == team2 ~ game_wins + wins_team2,
            TRUE ~ game_wins
          )) %>%
          mutate(game_losses = case_when(
            true_rank == team1 ~ game_losses + losses_team1,
            true_rank == team2 ~ game_losses + losses_team2,
            TRUE ~ game_losses
          ))
        }
        # Determine the match winner based on series results
        series_winner <- ifelse(wins_team1 > wins_team2, team1, team2)
        series_loser <- ifelse(series_winner == team1, team2, team1)
        cat(sprintf("Matchup:  %s vs %s | Winner: %s\n: | Loser: %s\n", team1, team2, series_winner, series_loser))
        next_winners <- c(next_winners, series_winner)
        round_losers <- c(round_losers, series_loser)
    }
     winners_bracket <- next_winners  # Winners move to the next round
     losers_bracket <- c(losers_bracket, round_losers)
     if(length(losers_bracket) %% 2 != 0){losers_bracket <- c(losers_bracket, NA)}

@@

    # Losers' bracket
    if (length(losers_bracket) > 0) {
      if(length(losers_bracket) > 1){
      cat("\n--- Loser's Bracket --- \n")
      }
      next_losers <- c()

@@

          # Determine overall series winner
          if (wins_team1 > series / 2 || wins_team2 > series / 2) {
            break
          }
        }
        if (!any(is.na(c(team1, team2)))) {
          cat(sprintf("Losses - %s: %d | %s: %d\n", team1, losses_team1, team2, losses_team2))
          permuted_df <- permuted_df %>%
          mutate(game_wins = case_when(
            true_rank == team1 ~ game_wins + wins_team1,
            true_rank == team2 ~ game_wins + wins_team2,
            TRUE ~ game_wins
          )) %>%
          mutate(game_losses = case_when(
            true_rank == team1 ~ game_losses + losses_team1,
            true_rank == team2 ~ game_losses + losses_team2,
            TRUE ~ game_losses
          ))
        }

        if(is.na(team1) && is.na(team2)){
          series_winner <- NA
          next_losers <- c(next_losers, NA)
        } else{
        series_winner <- ifelse(wins_team1 > wins_team2, team1, team2)
        series_loser <- ifelse(wins_team1 > wins_team2, team2, team1)
        next_losers <- c(next_losers, series_winner)
        eliminated <- c(eliminated, series_loser)
        cat(sprintf("Matchup:  %s vs %s | Winner: %s\n:", team1, team2, series_winner))
        }
      }
        team_rank <- team_rank - 1
        permuted_df$rank_hat[permuted_df$true_rank %in% eliminated] <- team_rank
        losers_bracket <- next_losers

@@

# Final matchup between winners and losers
    cat("\n--- Final Match ---\n")
    final_match <- c(winners_bracket, losers_bracket)
    if(length(final_match) == 2){
    team1 <- final_match[1]
    team2 <- final_match[2]
    strength1 <- strengths[which(permuted_df$true_rank == team1)]
    strength2 <- strengths[which(permuted_df$true_rank == team2)]

@@

      if (wins_team1 > series / 2 || wins_team2 > series / 2) {
        break
      }
    }
    if (!any(is.na(c(team1, team2)))) {
      cat(sprintf("Losses - %s: %d | %s: %d\n", team1, losses_team1, team2, losses_team2))
      permuted_df <- permuted_df %>%
        mutate(game_wins = case_when(
          true_rank == team1 ~ game_wins + wins_team1,
          true_rank == team2 ~ game_wins + wins_team2,
          TRUE ~ game_wins
        )) %>%
        mutate(game_losses = case_when(
          true_rank == team1 ~ game_losses + losses_team1,
          true_rank == team2 ~ game_losses + losses_team2,
          TRUE ~ game_losses
          ))
    }

    if (wins_team1 > wins_team2) {
      cat(sprintf("Matchup:  %s vs %s | Winner: %s\n:", team1, team2, team1))
      permuted_df$rank_hat[permuted_df$true_rank == team1] <- 1
      permuted_df$rank_hat[permuted_df$true_rank == team2] <- 2
    } else {
      cat(sprintf("Matchup:  %s vs %s | Winner: %s\n:", team1, team2, team2))
      # second game if the loser's bracket team wins
      for (game in 1:series) {
      match_winner <- simulate_match(team1, team2, strength1, strength2)
      if (match_winner == team1) {
        wins_team1 <- wins_team1 + 1
        losses_team2 <- losses_team2 + 1
      } else {
        wins_team2 <- wins_team2 + 1
        losses_team1 <- losses_team1 + 1
      }

      if (wins_team1 > series / 2 || wins_team2 > series / 2) {
        break
      }
    }
      if (!any(is.na(c(team1, team2)))) {
        cat(sprintf("Wins - %s: %d | %s: %d\n", team1, wins_team1, team2, wins_team2))
        permuted_df <- permuted_df %>%
          mutate(game_wins = case_when(
            true_rank == team1 ~ game_wins + wins_team1,
            true_rank == team2 ~ game_wins + wins_team2 - 1, # wins_team2 = 2 in loop but only need to add one game
            TRUE ~ game_wins
          )) %>%
          mutate(game_losses = case_when(
            true_rank == team1 ~ game_losses + losses_team1 - 1,
            true_rank == team2 ~ game_losses + losses_team2,
            TRUE ~ game_losses
          ))
      }
      if (wins_team1 > wins_team2) {
      cat(sprintf("Matchup:  %s vs %s | Winner: %s\n:", team1, team2, team1))
      permuted_df$rank_hat[permuted_df$true_rank == team1] <- 1
      permuted_df$rank_hat[permuted_df$true_rank == team2] <- 2
    } else {
      cat(sprintf("Matchup:  %s vs %s | Winner: %s\n:", team1, team2, team2))
      permuted_df$rank_hat[permuted_df$true_rank == team2] <- 1
      permuted_df$rank_hat[permuted_df$true_rank == team1] <- 2
        }
      }
    }
  }
}
    permuted_df$seed <- rep(paste(seeding_structure[i, ], collapse = ", "), nrow(permuted_df))
    results[[i]] <- permuted_df
}
  final_results <- do.call(rbind, results)

  # Remove ties if ties = FALSE
  if (!ties) {
    final_results <- final_results %>%
      filter(!is.na(true_rank)) %>%
      arrange(true_rank) %>%
      mutate(rank_hat = rank(rank_hat, ties.method = "random"))
  }
  else{
    final_results <- final_results %>%
      filter(!is.na(true_rank)) %>%
      arrange(rank_hat) %>%
      mutate(rank_hat = rank(rank_hat, ties.method = "average"))
  }

@@

test <- simulate_double_elimination_tournament(num_teams = 8, distribution = "Normal", ties=F, series=1)
#test
#manual_test <- simulate_double_elimination_tournament(num_teams = 4, distribution = "Manual", ties=F, series=1)
#manual_test
